---
output:
  html_document
bibliography: ref.bib
---

# Principal components analysis

```{r, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, error=FALSE)
```

## Motivation

We use principal components analysis (PCA) to clean up and compact the log-normalized expression matrix.
Consider each gene as a dimension of our dataset where the cells are the observations, i.e., each cell's expression profile defines its location in the high-dimensional expression space.
PCA discovers axes in this high-dimensional space that capture the largest amount of variation [@pearson1901lines].
Each principal component (PC) corresponds to an axis in this space, where the earliest PCs capture the dominant factors of heterogeneity in our data.
The idea is to use the first few PCs to approximate our original dataset^[
And in fact, PCA does provide the optimal low-rank approximation, based on its relation to the SVD and the Eckart-Young-Mirsky theorem.
This is arguably one of the rare cases of mathematical rigor in this entire book.].
Similarly, the Euclidean distances between cells in the PC space are approximations of the corresponding distances in the original dataset.
PCA effectively compresses multiple genes into a much smaller number of dimensions (a.k.a., "eigengenes", from @langfelder2007eigengene),
which allows us to use a much smaller matrix in downstream steps like clustering.

## Getting the top PCs

Our assumption is that biological processes affect multiple genes in a coordinated manner.
This means that the earlier PCs are likely to represent biological structure as more variation can be captured by considering the correlated behavior of many genes.
In contrast, random technical or biological noise is expected to affect each gene independently.
There is unlikely to be an axis that can capture random variation across many genes, meaning that noise should mostly be concentrated in the later PCs.
By retaining the earlier PCs, we can focus on the biological signal while removing random noise.
To demonstrate, we'll pull out our favorite mouse brain dataset from @zeisel2015brain:

```{r}
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()
is.mito.zeisel <- rowData(sce.zeisel)$featureType=="mito"

# Performing some QC to set up the dataset prior to normalization. 
library(scrapper)
sce.qc.zeisel <- quickRnaQc.se(sce.zeisel, subsets=list(MT=is.mito.zeisel), altexp.proportions="ERCC")
sce.qc.zeisel <- sce.qc.zeisel[,sce.qc.zeisel$keep]

# Computing log-normalized expression values.
sce.norm.zeisel <- normalizeRnaCounts.se(sce.qc.zeisel, size.factors=sce.qc.zeisel$sum)

# Computing the variances.
sce.var.zeisel <- chooseRnaHvgs.se(sce.norm.zeisel, more.var.args=list(use.min.width=TRUE))
```

We run the PCA on our HVG-filtered log-normalized expression matrix to compute the top 25 PCs.
This returns a matrix of "PC scores", i.e., the coordinates for each cell in the new low-dimensional space.
As discussed in Chapter \@ref(feature-selection), we restrict this step to the top HVGs to reduce the impact of random noise. 
While PCA is robust to noise, too much of it may cause the earlier PCs to ignore meaningful structure [@johnstone2009consistency].

```{r}
hvgs.zeisel <- rowData(sce.var.zeisel)$hvg
summary(hvgs.zeisel)

sce.pca.zeisel <- runPca.se(sce.var.zeisel, features=hvgs.zeisel, number=25)
dim(reducedDim(sce.pca.zeisel, "PCA"))
```

We then use the PC score matrix in place of the original log-expression matrix in downstream steps like clustering (Chapter \@ref(clustering)) and visualization (Chapter \@ref(visualization)). 
The former is much smaller than the latter as the relevant heterogeneity is compacted into a smaller number of dimensions.
This enables more efficient calculation of distances between cells and reduces the impact of random noise.

```{r}
# This performs graph-based clustering on the PC score matrix. See the linked
# chapters for more in-depth discussion of this step.
table(clusterGraph.se(sce.pca.zeisel, reddim.type="PCA")$clusters)
```

## How many PCs?

The million dollar question is, how many of the top PCs should we retain for downstream analyses?
Using more PCs will retain more biological signal at the cost of including more unintersting noise that might mask that signal.
As with the number of HVGs, it is hard to determine whether an "optimal" choice exists here.
Sure, technical variation is almost always uninteresting,
but there is no straightforward way to automatically determine which aspects of biological variation are relevant to a particular scientific question.
For example, heterogeneity within a population might be interesting when studying continuous processes like metabolic flux or differentiation potential,
but could be considered noise in applications that only aim to distinguish between distinct cell types.

Most practitioners will simply set $d$ to a "reasonable" but arbitrary value, typically ranging from 10 to 50.
This is often satisfactory as the later PCs explain so little variance that their inclusion or omission has no major effect.
For example, in the Zeisel dataset, very few PCs explain more than 1\% of the variance in the entire dataset (Figure \@ref(fig:zeisel-scree)).
Choosing between, say, 20 and 40 PCs would not even amount to 5 percentage points' worth of difference in variance.
In fact, the main consequence of using more PCs is simply an increase in compute time as more dimensions need to be processed, 
but most PC-related calculations are fast enough that this is not a practical concern.

```{r zeisel-scree, fig.cap="Percentage of variance explained by successive PCs in the Zeisel dataset, shown on a log-scale."}
sce.more.zeisel <- runPca.se(sce.var.zeisel, features=hvgs.zeisel, number=50)
pca.meta <- metadata(sce.more.zeisel)$PCA
percent.var <- pca.meta$variance.explained / pca.meta$total.variance * 100
plot(percent.var, log="y", xlab="PC", ylab="Variance explained (%)", type="b")
```

```{r, echo=FALSE}
stopifnot(sum(percent.var[20:40]) < 5)
```

If we really must try to guess the "best" number of PCs^[Probably because a reviewer asked us to.], here are a few approaches:

- We can choose the elbow point in the scree plot (Figure \@ref(fig:zeisel-scree)), e.g., using the `findElbowPoint()` function from the `r Biocpkg("PCAtools")` package.
  The assumption is that there should be a sharp drop in the percentage of variance explained when we move past the last PC corresponding to biological structure.
  However, the ideal cut-off can be difficult to gauge when there are sources of weaker biological variation. 
- We can keep the number of PCs that cumulatively explain variance equal to the sum of the biological components among the HVGs.
  This relies on the decomposition of each gene's variance into biological and technical components (see Chapter \@ref(feature-selection)).
  In practice, the separation between biological and technical variation is usually not so clear-cut;
  some technical noise will still contribute to the earlier PCs while some biological variation will still be present in the later PCs.
- We can use random matrix theory to select an appropriate number of PCs.
  This might involve the Marchenko-Pastur limit [@shekhar2016comprehensive], Horn's parallel analysis [@horn1965rationale],
  or the Gavish-Donoho threshold for optimal reconstruction [@gavish2014optimal] (see relevant functions in `r Biocpkg("PCAtools")`).
  Each of these methods has its own limitations, e.g., requirement for i.i.d. noise.

But if we're really concerned about the number of PCs, it's probably just better to repeat the analysis with different number of PCs.
This allows us explore other perspectives of the data at different trade-offs between biological signal and technical noise.

## Blocking on uninteresting factors {#pca-block}

Larger datasets typically consist of multiple blocks of cells with uninteresting differences between them, e.g., batch effects, variability between donors.
We don't want to waste our top PCs on capturing these differences - instead, we want our PCA to focus on the biological structure within each block.
To demonstrate, let's look at two 10X Genomics datasets of human PBMCs [@zheng2017massively]. 
These datasets are effectively replicates so any differences between them are uninteresting and should be ignored. 

```{r}
library(TENxPBMCData)
sce.pbmc3k <- TENxPBMCData('pbmc3k')
sce.pbmc4k <- TENxPBMCData('pbmc4k')

# Finding a common set of genes across all datasets to allow us to combine
# everything into a single object. This is only necessary if the different
# datasets were processed with different genome annotations.
inter.pbmc <- intersect(rownames(sce.pbmc3k), rownames(sce.pbmc4k))
sce.pbmc <- combineCols(sce.pbmc3k[inter.pbmc,], sce.pbmc4k[inter.pbmc,])
sce.pbmc$block <- rep(c("3k", "4k"), c(ncol(sce.pbmc3k), ncol(sce.pbmc4k)))

# Loading the counts into memory. for convenience.
counts(sce.pbmc) <- as(counts(sce.pbmc), "dgCMatrix")

# Quality control, blocking on the dataset of origin for each cell.
is.mito.pbmc <- grep("MT", rowData(sce.pbmc)$Symbol)
library(scrapper)
sce.qc.pbmc <- quickRnaQc.se(
    sce.pbmc,
    subsets=list(MT=is.mito.pbmc),
    block=sce.pbmc$block
)
sce.qc.pbmc <- sce.pbmc[,sce.qc.pbmc$keep]

# Normalization, blocking on the dataset of origin for each cell.
sce.norm.pbmc <- normalizeRnaCounts.se(
    sce.qc.pbmc,
    size.factors=sce.qc.pbmc$sum,
    block=sce.qc.pbmc$block
)

# We now choose the top HVGs, with blocking.
sce.var.pbmc <- chooseRnaHvgs.se(
    sce.norm.pbmc, 
    block=sce.norm.pbmc$block
)
```

We specify `block=` to instruct `runPca.se()` to focus on the variation within each block.
Roughly speaking, this is equivalent to computing the gene-gene covariances within each block,
taking the weighted average across blocks, and using the subsequent covariance matrix to define the axes of variation.
This reduces the contribution of uninteresting differences between blocks and provides more opportunities for the top PCs to capture biological variation within each block. 
For example, blocking removes some of the shift between datasets in the first two PCs (Figure \@ref(fig:pca-pbmc)).

```{r pca-pbmc, fig.width=10, fig.height=5, fig.cap="First two PCs for the combined PBMC datasets, before and after blocking on the dataset of origin. Each point represents a cell, colored by its dataset."}
sce.block.pbmc <- runPca.se(
    sce.var.pbmc,
    features=rowData(sce.var.pbmc)$hvg,
    block=sce.var.pbmc$block,
    number=10
)

# For comparison, let's look at the unblocked PCA.
sce.pca.pbmc <- runPca.se(
    sce.var.pbmc,
    features=rowData(sce.var.pbmc)$hvg,
    number=10
)

library(scater)
gridExtra::grid.arrange(
    plotReducedDim(sce.pca.pbmc, dimred="PCA", colour_by="block") +
        ggtitle("Without blocking"),
    plotReducedDim(sce.block.pbmc, dimred="PCA", colour_by="block") +
        ggtitle("Blocked"),
    ncol=2
)
```

It's worth mentioning that blocking in `runPca.se()` is not guaranteed to remove the differences between blocks.
If we're lucky, blocking alone is sufficient if all of the inter-block differences are orthogonal to the major biological variation within the blocks,
such that taking the first few PCs will focus on the latter and remove the former.
In practice, inter-block differences tend to have some biological component that will be preserved within the first few PCs.
For example, imagine our two PBMC datasets were derived from different patients who naturally varied in their baseline CD3 expression.
This variation between patients would probably be captured in the top PCs given that at least one PC would be involved in separating T cells from other types.
If such differences are uninteresting, their removal requires some additional effort (see Chapter \@ref(batch-correction)) prior to downstream steps like clustering.
Nonetheless, blocking is still helpful as it eliminates at least some of these differences and preserves more biological signal in the top PCs^[
Some folks like to go through their PCs and remove those that are correlated with uninteresting factors like the batch of origin.
This tends to be a dangerous exercise as you may well be removing relevant biological differences that also happen to be correlated with the batch.
It would be better to block within the PCA itself to avoid wasting top PCs on batch effects in the first place.].

For completeness' sake: we could attempt to remove the differences between blocks explicitly by setting `center.scores.by.block=TRUE`.
This is equivalent to performing a linear regression on the expression values with respect to the block and deriving the PC scores from the residuals.
Regressing out the block removes any systematic difference in location between blocks but is only correct in very limiting circumstances,
e.g., assuming all blocks have the same subpopulation composition and the difference between blocks is consistent for all cell subpopulations.
Such assumptions may be appropriate in some situations - say, when technical replicates are involved - but are not generally applicable.
Don't be fooled by Figure \@ref(fig:pca-residuals-416b); the PBMC datasets are replicates and most other applications of `center.scores.by.block=TRUE` will not look as good.

```{r pca-residuals-pbmc, fig.cap="First two PCs for the combined PBMC datasets after blocking on the dataset of origin, where PC scores are computed from residuals. Each point represents a cell, colored by its dataset."}
sce.resid.pbmc <- runPca.se(
    sce.var.pbmc,
    features=rowData(sce.var.pbmc)$hvg,
    block=sce.var.pbmc$block,
    more.pca.args=list(center.scores.by.block=TRUE),
    number=10
)

library(scater)
plotReducedDim(sce.resid.pbmc, dimred="PCA", colour_by="block")
```

As with HVGs, we should only use `block=` for experimental factors that are not interesting.
If we were interested in the effects of oncogene induction, we should not block on it to ensure that the PCA can capture the associated changes in expression.
Sometimes, though, it is not obvious whether something is "interesting" or not,
as we may wish to ignore some biological differences to obtain a consistent set of clusters across treatment conditions, tissues, etc.
Check out Chapter \@ref(batch-correction) for a more detailed discussion.

## Visualizing the PCs

We might as well touch on another common use of PCA, which is visualization of high-dimensional data.
This is used in a variety of fields and applications including bulk RNA-seq, but is not as effective for scRNA-seq data. 
If we're lucky, our population structure is simple enough that the first two PCs capture most of the relevant biology. 
However, in most cases, relevant biological heterogeneity is spread throughout 10-50 PCs that are much harder to visualize.
For example, examination of the top 4 PCs is still insufficient to resolve all subpopulations identified by @zeisel2015brain (Figure \@ref(fig:zeisel-pca-multi)).

```{r zeisel-pca-multi, fig.height=10, fig.width=10, fig.cap="PCA plot of the first 4 PCs in the Zeisel brain data. Each point is a cell, coloured according to the annotation provided by the original authors."}
library(scater)
plotReducedDim(
    sce.pca.zeisel,
    dimred="PCA",
    ncomponents=4,
    colour_by="level1class"
)
```

The problem here is that PCA is a linear technique, i.e., only variation along a line in high-dimensional space is captured by each PC.
As such, it cannot efficiently represent high-dimensional differences in the first 2 PCs.
If the first PC is devoted to resolving the biggest difference between subpopulations, and the second PC is devoted to resolving the next biggest difference,
then the remaining differences will not be visible in the plot.
That said, PCA is still useful as the top PCs are often used as input to more sophisticated algorithms for dimensionality reduction (Chapter \@ref(visualization)).

## Session information {-}

```{r}
sessionInfo()
```
